---
title: "Covid-19 Vaccination Shiny App"
author: "Anna Sikora, Alparslan Erol, Tesfahun Tegene Boshe"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Introduction

This markdown is created for the purpose of initial data analysis before creating an app using shiny. The purpose of the app is to create a visualization of the vaccination process around the world and the track progress. 

The data used for the analysis comes from git hub page provided below:
https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations

The data consists of 12 columns:

* date
 + location
 + iso_code
 + total_vaccinations
 + people_vaccinated
 + people_fully_vaccinated
 + daily_vaccinations_raw
 + daily_vaccinations
 + total_vaccinations_per_hundred
 + people_vaccinated_per_hundred
 + people_fully_vaccinated_per_hundred 
 + daily_vaccinations_per_million
 
### Exploratory Data Analysis
#### Importing neccessary packages and removing scientific notation
```{r message=FALSE, warning=FALSE}
library(rio)
library(e1071)   
library(data.table)
library(ggplot2)
library(dplyr)

options(scipen=999)
```

#### Importing the data from github and loading 

```{r}
vaccinations <- data.table(import("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/vaccinations.csv"))
```

### Basic Data analysis

#### Summary statistics

```{r}
summary(vaccinations)
```

#### Column data types

```{r}
str(vaccinations)
```

#### Removing NA's

```{r}
vaccinations <- na.omit(vaccinations)
```

#### Exploring the regions in the dataset

As we can see, the world data is divided according to various criteria:

* data for the whole world
* division by continent
* division by income level
* division by country

```{r}
unique(vaccinations$location)
```


#### World results

```{r}
world <- vaccinations[location == "World"]
```
```{r fig.width=8, fig.height=4}
w0<-ggplot(data = world) +
  geom_line(aes(x=date, y=total_vaccinations),color = "blueviolet", linetype = "solid", size = 1)+
  xlab("Date")+
  ylab("Total vaccinations")+
  labs(title = "Total vaccinations evolution across time")+
  theme_update(plot.title = element_text(hjust=0.5))
w0
```


```{r fig.width=8, fig.height=4}
w1<-ggplot(data = world) +
  geom_line(aes(x=date, y=people_vaccinated),color = "grey39", linetype = "solid", size = 1)+
  geom_line(aes(x=date, y=people_fully_vaccinated),color = "green2", linetype = "solid", size = 1)+
  xlab("Date")+
  ylab("Vaccinations")+
  labs(title = "People vaccinated vs. people fully vaccinated")+
  theme_update(plot.title = element_text(hjust=0.5))
w1
```

```{r fig.width=8, fig.height=4}
w2<-ggplot(data = world) +
  geom_line(aes(x=date, y=daily_vaccinations_per_million),color = "navyblue", linetype = "solid", size = 1)+
  xlab("Date")+
  ylab("Vaccinations")+
  labs(title = "Daily vaccinations per million around the world")+
  theme_update(plot.title = element_text(hjust=0.5))
w2
 
```
### Results by countries 

```{r, warning=FALSE}
vaccinations <- vaccinations %>%
  select(location, total_vaccinations, iso_code) %>%
  filter(!location %in% c("World","Europe", "Low income", "Lower middle income", "Africa", "Asia", "European Union", "High income", "Oceania", "South America", "Upper middle income", "Australia", "North America"))

data <-vaccinations[, sum(total_vaccinations), by = location]
```
```{r fig.width=8, fig.height=12}
options(repr.plot.width=12,repr.plot.height=50)
data %>%
  ggplot(aes(x = V1, y = location)) +
  geom_bar(aes(fill = location), position = "identity", stat = "identity", show.legend = FALSE) +
  labs(x = "Total Vaccinations", y = "Country") +
  labs(title = "Total vaccinations by country")+
  theme(axis.text.y = element_text(size=6),panel.grid = element_blank())
```

### Grouping by the country's income

```{r, warning=FALSE}
vaccinations <- data.table(import("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/vaccinations.csv"))

vaccinations <- na.omit(vaccinations)

income <- vaccinations %>%
  select(location, total_vaccinations, iso_code) %>%
  filter(location %in% c("Low income", "Lower middle income","High income", "Upper middle income"))
```

```{r fig.width=8, fig.height=5, warning=FALSE}
options(repr.plot.width=16,repr.plot.height=12)
income %>%
  ggplot(aes(x = location, y = total_vaccinations)) +
  geom_bar(aes(fill = total_vaccinations), position = "identity", stat = "identity", show.legend = FALSE) +
  labs(x = "Income Groups", y = "Total Vaccinations") +
  theme(panel.grid = element_blank(), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```


### Total cases by continents
```{r , warning=FALSE}
continents <- vaccinations %>%
  select(location, total_vaccinations, iso_code) %>%
  filter(location %in% c("Europe","Africa", "Asia", "Oceania", "South America","Australia", "North America"))
```

```{r fig.width=8, fig.height=5, warning=FALSE}
continents <-continents[, sum(total_vaccinations), by = location]
continents <- na.omit(continents)

slices <- continents$V1
labls <- continents$location

pie(slices,labels=labls,explode=0.1, col=rainbow(length(labls)),
      main="Total vaccinations by continent ")
```